---
title: "fefifo_metabolomics"
author: "HCW"
date: "11/21/2020"
output: html_document
---

```{r}
metab_data_raw <- read.csv("~/R/metabolomics_projects/serum_c18_pos_neg_norm_data.csv")

save_path_16S <- "../../data/16S/"
samdf_diversity <- readRDS(file = paste(save_path_16S, "samdf_diversity.rds", sep = "")) 
ps_dada2_with_tree <- readRDS( file = paste(save_path_16S, "phyloseq_obj_PilotStudy_tree_redo_fwdONLY.rds", sep = ""))
df_ranks_gather <- readRDS(file = paste(save_path_16S, "asv_rank_fiber_fermented.rds", sep = ""))
save_path_figures <- "../../plots/"


save_path_olink <- "../../data/Olink/cleaned/"
olink_data <- read_csv(paste(save_path_olink, "olink_data_cleaned.csv", sep = "")) %>% mutate(Timepoint = as.numeric(Timepoint))

metab_data <- metab_data_raw %>% 
  filter(Study=="Pilot") %>% 
  filter(! Participant %in% c(8000,8012) & Timepoint !=5) %>% 
  mutate(Timepoint_label=ifelse(Timepoint %in% c(1,2),"Baseline","Maint"),
         Timepoint=as.factor(Timepoint)) %>% 
  gather(key=metab,value=metab_value,-X,-Tube_ID,-Sample_type,-Study,-Participant,-Timepoint,-Timepoint_label) %>%
  left_join(unique(select(samdf_diversity,Participant,Group))) %>% 
  select(Participant,Group,Timepoint,Timepoint_label,metab,metab_value)

metab_data_spread <- metab_data %>% 
  spread(key=metab,value=metab_value)

```


```{r}
paired_df_ppt <- function(df,tp_1,tp_2){
  df <- df %>% 
    arrange(Participant)
  temp_1 <- df %>% 
    filter(Timepoint_label==tp_1) 
  temp_2 <- df %>% 
    filter(Timepoint_label==tp_2)
  ppt_list <- intersect(temp_1$Participant,temp_2$Participant)
  
  df_paired <- bind_rows(
    filter(temp_1, Participant %in% ppt_list),
    filter(temp_2,Participant %in% ppt_list)
  )
  return(df_paired)
}

var_function <- function(df_spread,num_quantile){
  vars_vector <- df_spread %>% 
    select(-Participant,-Timepoint,-Group,-Timepoint_label) %>% 
    as.matrix() %>% 
    colVars()
  name_vector <- df_spread %>% 
    select(-Participant,-Timepoint,-Group,-Timepoint_label) %>% 
    colnames()
  filt_names <- name_vector[vars_vector>quantile(vars_vector)[num_quantile]] #top 75% varying proteins =2
  return(filt_names)
}
```

Paired siggenes end of intervention to baseline
-filtered to top 75% most varying metabolites (like the host proteins)
-no sig individual metabolites from baseline to end of intervention
```{r}
metab_data_fermented <- metab_data %>% 
  filter(Group=="Fermented") %>% 
  spread(key=metab,value=metab_value)

metab_fermented_names <- var_function(metab_data_fermented,2)
metab_data_fermented_filt <- metab_data %>% 
  filter(Group=="Fermented" & metab %in% metab_fermented_names) %>% 
  spread(key=metab,value=metab_value)

metab_data_fermented_paired <- paired_df_ppt(metab_data_fermented_filt,"Baseline","Maint")
set.seed(20)
siggenes_cl <- c((-1:-18), 1:18)
siggenes_data <- t(metab_data_fermented_paired[-c(1:4)])
siggenes_output <- siggenes::sam(data=siggenes_data,cl=siggenes_cl)
summary(siggenes_output)
delta_siggenes <- findDelta(siggenes_output, fdr = 0.05)

#fiber group
metab_data_fiber <- metab_data %>% 
  filter(Group=="Fiber") %>% 
  spread(key=metab,value=metab_value)

metab_fiber_names <- var_function(metab_data_fiber,2)
metab_data_fiber_filt <- metab_data %>% 
  filter(Group=="Fiber" & metab %in% metab_fiber_names) %>% 
  spread(key=metab,value=metab_value)

metab_data_fiber_paired <- paired_df_ppt(metab_data_fiber_filt,"Baseline","Maint")
set.seed(20)
siggenes_cl <- c((-1:-16), 1:16)
siggenes_data <- t(metab_data_fiber_paired[-c(1:4)])
siggenes_output <- siggenes::sam(data=siggenes_data,cl=siggenes_cl)
summary(siggenes_output)
delta_siggenes <- findDelta(siggenes_output, fdr = 0.05)
```

Beta diversity of metabolites from baseline to end of maint
-no sig. difference in clusters between timepoints
```{r}
metab_data_fermented_paired_raw <- metab_data_fermented_paired %>% 
  select(-c(Participant:Timepoint_label))
permanova <- adonis(metab_data_fermented_paired_raw ~ Timepoint_label, data = metab_data_fermented_paired, method='euclidean',na.rm = TRUE)
as.data.frame(permanova$aov.tab)["Timepoint_label", "Pr(>F)"]

metab_data_fiber_paired_raw <- metab_data_fiber_paired %>% 
  select(-c(Participant:Timepoint_label))
permanova <- adonis(metab_data_fiber_paired_raw ~ Timepoint_label, data = metab_data_fiber_paired, method='euclidean',na.rm = TRUE)
as.data.frame(permanova$aov.tab)["Timepoint_label", "Pr(>F)"]
```
PCA on metabolites
```{r}
pca_fermented <- prcomp(metab_data_fermented_paired_raw,scale=TRUE)
pca_fermented_df <- pca_fermented[["x"]] %>% as.data.frame %>% 
  cbind(select(metab_data_fermented_paired,Participant,Group,Timepoint_label),.)

ggplot(pca_fermented_df,aes(x=PC1,y=PC2,colour=Timepoint_label))+
  geom_point()
```


Calculate metabolite difference across time for each participant
-PCA by group, no sig. difference in clustering (small trend)
```{r}
metab_data_diff <- metab_data %>% 
  # filter(Group %in% c("Fiber","Fermented")) %>% 
  # gather(key=metab,value=metab_value,-c(Participant:Timepoint_label)) %>% 
  select(-Timepoint) %>% 
  spread(key=Timepoint_label,value=metab_value) %>% 
  mutate(diff=Maint-Baseline)

var_function <- function(df_spread,num_quantile){
  vars_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    as.matrix() %>% 
    colVars()
  name_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    colnames()
  filt_names <- name_vector[vars_vector>quantile(vars_vector)[num_quantile]] #top 75% varying proteins =2
  return(filt_names)
}

metab_diff_names <- metab_data_diff %>% 
  select(-Baseline,-Maint) %>% 
  spread(key=metab,value=diff) %>%
  na.omit() %>% 
  var_function(.,2)

metab_data_filt_diff_spread <- metab_data_diff %>% 
  filter(metab %in% metab_diff_names) %>%
  select(-Baseline,-Maint) %>% 
  spread(key=metab,value=diff) %>% 
  na.omit()

pca_diff <- prcomp(select(metab_data_filt_diff_spread,-Participant,-Group),scale=TRUE)
pca_diff_df <- pca_diff[["x"]] %>% as.data.frame %>% 
  cbind(select(metab_data_filt_diff_spread,Participant,Group),.)

ggplot(pca_diff_df,aes(x=PC1,y=PC2,colour=Group))+
  geom_point()

pca_diff_df_raw <- pca_diff_df %>% 
  select(-c(Participant:Group)) 
permanova <- adonis(pca_diff_df_raw ~ Group, data = pca_diff_df, method='eu',na.rm = TRUE)
as.data.frame(permanova$aov.tab)["Group", "Pr(>F)"]
```

Correlate alpha diversity with each metabolite
```{r}
lme1by1HCW <- function(df,id, vectorNames, yVar,xColName){
  pValueList <- c()
  xList <- c()
  correValueList <- c()
  stdEList <-c()
  for (i in vectorNames){
    # df=df_raw
    # vectorNames=taxa_names
    # i=vectorNames[1]
    newdf <- data.frame(yVar = df[,yVar], xColName = df[, i], id = df[,id])
    #print(newdf)
    lmeData <- lme(yVar ~ xColName, data = newdf,random = ~1|id,na.action = na.omit)
    pVal = summary(lmeData)$tTable[2,5]
    corre = summary(lmeData)$tTable[2,1]
    stdE = summary(lmeData)$tTable[2,2]
    xList <- c(xList,i)
    pValueList <- c(pValueList,pVal)
    correValueList <- c(correValueList,corre)
    stdEList <- c(stdEList,stdE)
  }
  pTable <- c()
  pTable <- data.frame(xColName = xList, correlation = correValueList, pValue = pValueList, pValueAdj = p.adjust(pValueList,method="BH"),stdE=stdEList)
}


var_function <- function(df_spread,num_quantile){
  vars_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    as.matrix() %>% 
    colVars()
  name_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    colnames()
  filt_names <- name_vector[vars_vector>quantile(vars_vector)[num_quantile]] #top 75% varying proteins =2
  return(filt_names)
}
```

```{r}
samdf_diversity_tp2 <- samdf_diversity %>% 
  filter(Timepoint==2) %>% 
  select(Participant,Group,Observed) %>% 
  dplyr::rename(Observed_2=Observed)

metab_data_alpha_div <- metab_data %>% 
  left_join(.,select(samdf_diversity,Participant,Group,Timepoint,Observed)) %>% 
  select(Participant,Group,Timepoint,Timepoint_label,Observed,metab,metab_value) %>% 
  left_join(.,samdf_diversity_tp2) %>% 
  mutate(Observed_corrected=ifelse(is.na(Observed) & Timepoint_label=="Baseline",Observed_2,Observed)) #if participant does not have an observed vlaue for tp=1, replace with tp2=2. Not doing this for missing maintenance tp=7

metab_data_alpha_div_fermented_filt <- metab_data_alpha_div %>% 
  filter(Group=="Fermented" & metab %in% metab_fermented_names) %>% 
  spread(key=metab,value=metab_value) %>% 
  na.omit()
metab_ferm_lme <- lme1by1HCW(metab_data_alpha_div_fermented_filt,"Participant",metab_fermented_names,"Observed_corrected",metab)

```
Correlate alpha diversity with PC space
```{r}
pca_fermented_df_alpha_div <- inner_join(unique(select(metab_data_alpha_div,Participant,Group,Timepoint_label,Observed_corrected)),
                                         pca_fermented_df) 
lme_obj_pc1 <- lme(Observed_corrected ~ PC1, data = pca_fermented_df_alpha_div,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc1)$tTable
lme_obj_pc2 <- lme(Observed_corrected ~ PC2, data = pca_fermented_df_alpha_div,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc2)$tTable
lme_obj_pc3 <- lme(Observed_corrected ~ PC3, data = pca_fermented_df_alpha_div,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc3)$tTable

ggplot(pca_fermented_df_alpha_div,aes(x=PC2,y=Observed_corrected,colour=Timepoint_label))+
  geom_point()

pca_fermented_df_alpha_div_diff <- pca_fermented_df_alpha_div %>% 
  gather(key=Param, value=Value,-Participant,-Group,-Timepoint_label) %>% 
  spread(key=Timepoint_label,value=Value) %>% 
  mutate(Difference=Maint-Baseline) %>% 
  select(-Baseline,-Maint) %>% 
  spread(key=Param,value=Difference)

lme_obj_pc1 <- lme(Observed_corrected ~ PC1, data = pca_fermented_df_alpha_div_diff,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc1)$tTable
lme_obj_pc2 <- lme(Observed_corrected ~ PC2, data = pca_fermented_df_alpha_div_diff,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc2)$tTable
lme_obj_pc3 <- lme(Observed_corrected ~ PC3, data = pca_fermented_df_alpha_div_diff,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc3)$tTable

fermented_intercept=summary(lme_obj_pc2)$tTable[1,1]
fermented_slope=summary(lme_obj_pc2)$tTable[2,1]
ggplot(pca_fermented_df_alpha_div_diff,aes(x=PC2,y=Observed_corrected))+
  geom_point()+
  theme_bw()+
  theme(text = element_text(size=18))+
  xlab("PC2 for metabolite difference")+
  ylab("Observed ASV difference")+
  geom_abline(intercept=fermented_intercept,slope=fermented_slope)
#ggsave(paste(save_path_figures,"pca_metabs_alpha_div_fermented.pdf"),width = 6,height=4)

#plot loadings of pc2
pca_fermented_loadings <- pca_fermented[["rotation"]] %>% as.data.frame 

pc2_fermented_loadings <- pca_fermented_loadings %>% 
  mutate(metab=row.names(.),
         PC2_mag=abs(PC2)) %>% 
  select(metab,PC2,PC2_mag) %>% 
  arrange(-PC2)
#cleaned names in excel
save_path_metabolomics <- "../../data/metabolomics/"
metab_pc2_clean <- read.csv(file = paste(save_path_metabolomics, "pc2_loadings_metab_clean_names.csv", sep = "")) %>% 
  mutate(sign=ifelse(PC2.loading>0,"pos","neg"))

ggplot(metab_pc2_clean,aes(x=reorder(Metabolite,PC2.loading),y=PC2.loading,fill=sign))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c("#2166ac","#b2182b"))+
  coord_flip()+
  theme_bw()+
  ylab("PC2 loading")+
  xlab("Metabolite")+
  theme(legend.position = "none")
ggsave(paste(save_path_figures,"pc2_loadings_alpha_div_fermented.pdf"),width = 6,height=4)

sum(pca_fermented_loadings$PC2^2)
```

Correlate metabolites with cytokines
```{r}
olink_data_gather <- olink_data %>% 
  gather(key=cytokine,value=cytokine_value,-Group,-Participant,-Timepoint) 
olink_data_fermented <- olink_data_gather %>% 
  filter(Group=="Fermented") %>% 
  mutate(cytokine=make.names(cytokine),
         Timepoint=as.factor(Timepoint)) %>% 
  spread(key=cytokine,value=cytokine_value) 


cytokine_list <- unique(olink_data_gather$cytokine) %>% make.names
cytokine_serum_metab_lme <- data.frame(cytokine=c(),
                                   xColName=c(),
                                   correlation=c(),
                                   pValue=c(),
                                   pValueAdj=c(),
                                   stdE=c())
for (i in cytokine_list){
  # i=cytokine_list[1]
  df_cytokine_metab <- inner_join(select(olink_data_fermented,Participant,Timepoint,all_of(i)),
                                  select(metab_data_fermented,-Timepoint_label)) %>% 
    na.omit() %>% 
    data.frame
  df_raw <- df_cytokine_metab %>% 
    select(-Participant,-Timepoint,-Group)
  df_raw <- df_raw[,colSums(df_raw) > 0]
  df_in <- df_cytokine_metab %>% select(Participant,colnames(df_raw))
  metab_names <- select(df_raw,-all_of(i)) %>% colnames
  metab_all_LME <- lme1by1HCW(df_in,"Participant",metab_names,i,metab) %>% mutate(cytokine=i)
  sig_metab <- metab_all_LME %>% filter(pValueAdj <= 0.05) 
  cytokine_serum_metab_lme <- bind_rows(cytokine_serum_metab_lme,sig_metab)
}

#calculate LME based on difference between baseline and maint for each participant
metab_data_fermented_diff <- metab_data_fermented %>% 
  gather(key=metab,value=metab_value,-Participant,-Group,-Timepoint,-Timepoint_label) %>% 
  select(-Timepoint) %>% 
  spread(key=Timepoint_label,value=metab_value) %>% 
  mutate(Diff=Maint-Baseline) %>% 
  select(-Baseline,-Maint) %>% 
  spread(key = metab, value=Diff)

olink_data_fermented_diff <- olink_data_fermented %>% 
  filter(Timepoint %in% c(1,6)) %>% 
  gather(key=cytokine,value=cytokine_value,-Participant,-Group,-Timepoint) %>% 
  spread(key=Timepoint,value=cytokine_value) %>% 
  mutate(Diff=`6`-`1`,
         cytokine=make.names(cytokine)) %>% 
  select(-`6`,-`1`) %>% 
  spread(key = cytokine, value=Diff)

cytokine_list <- select(olink_data_fermented_diff,-Participant,-Group) %>%colnames()
cytokine_serum_metab_diff_lme <- data.frame(cytokine=c(),
                                   xColName=c(),
                                   correlation=c(),
                                   pValue=c(),
                                   pValueAdj=c(),
                                   stdE=c())
for (i in cytokine_list){
  df_cytokine_metab <- inner_join(select(olink_data_fermented_diff,Participant,all_of(i)),
                                  metab_data_fermented_diff) %>% 
    na.omit() %>% 
    data.frame
  df_raw <- df_cytokine_metab %>% 
    select(-Participant,-Group) 
  df_raw <- df_raw[,colSums(df_raw) != 0]
  df_in <- df_cytokine_metab %>% select(Participant,colnames(df_raw)) %>% scale
  metab_names <- select(df_raw,-all_of(i)) %>% colnames
  metab_all_LME <- lme1by1HCW(df_in,"Participant",metab_names,i,metab) %>% mutate(cytokine=i)
  sig_metab <- metab_all_LME %>% filter(pValueAdj <= 0.05) 
  cytokine_serum_metab_diff_lme <- bind_rows(cytokine_serum_metab_diff_lme,sig_metab)
}

cytokine_length_lme <- table(cytokine_serum_metab_diff_lme$cytokine) %>% as.data.frame %>% 
  dplyr::rename(cytokine=Var1)
cytokine_serum_metab_diff_lme_plot <- cytokine_serum_metab_diff_lme %>% 
  mutate(log_pvalue=-log(pValueAdj)*round(correlation)) 

ggplot(cytokine_serum_metab_diff_lme_plot,aes(x=cytokine, y=reorder(xColName,-abs(log_pvalue)), colour = log_pvalue,size = abs(log_pvalue))) +
  geom_point()+
  theme_bw()+
  scale_colour_gradient2(low="red",mid="white",high="blue")+
  scale_size(range = c(5, 10), name="-log(p-value adj.)")

#remove duplicate metabolites from different ion modes of mass spec
metab_dupes <- c("glycocholic.acid_10140_c18neg","Glycochenodeoxycholic.acid.sodium.salt_22833540_c18pos","X4.Acetamidophenol_1983_c18neg","sebacic_5192_peak1_c18neg","Suberic.acid_10457_c18neg","FLAG_glycochenodeoxycholic.acid_12544_glycodeoxycholic.acid_3035026_c18pos")
cytokine_serum_metab_diff_lme_plot_clean <- cytokine_serum_metab_diff_lme_plot %>%
  filter(! xColName %in% metab_dupes)
#cleaned names in excel
save_path_metabolomics <- "../../data/metabolomics/"
metab_cytokine_lme_clean <- read.csv(file = paste(save_path_metabolomics, "metab_cytokine_clean_names.csv", sep = "")) 

ggplot(metab_cytokine_lme_clean,aes(x=cytokine, y=reorder(Metabolite_clean,-abs(log_pvalue)), colour = correlation>0,size = abs(log_pvalue))) +
  geom_point()+
  theme_bw()+
  scale_colour_manual(values=c("#2166ac","#b2182b"))+
  # scale_colour_gradient2(low="red",mid="white",high="blue")+
  scale_size(range = c(4, 10), name="-log(p-value adj.)")+
  xlab("Inflammatory Cytokine")+
  ylab("Metabolite")+
  theme(text = element_text(size=20))
#ggsave(paste(save_path_figures,"metab_cytokine_sig_LME.pdf"),width = 20,height=9)

```