---
title: "fefifo_metabolomics"
author: "HCW"
date: "11/21/2020"
output: html_document
---

```{r}
metab_data_raw <- read.csv("~/R/metabolomics_projects/serum_c18_pos_neg_norm_data.csv")

save_path_16S <- "../../data/16S/"
samdf_diversity <- readRDS(file = paste(save_path_16S, "samdf_diversity.rds", sep = "")) 
ps_dada2_with_tree <- readRDS( file = paste(save_path_16S, "phyloseq_obj_PilotStudy_tree_redo_fwdONLY.rds", sep = ""))
df_ranks_gather <- readRDS(file = paste(save_path_16S, "asv_rank_fiber_fermented.rds", sep = ""))
save_path_figures <- "../../plots/"


save_path_olink <- "../../data/Olink/cleaned/"
olink_data <- read_csv(paste(save_path_olink, "olink_data_cleaned.csv", sep = "")) %>% mutate(Timepoint = as.numeric(Timepoint))

metab_data <- metab_data_raw %>% 
  filter(Study=="Pilot") %>% 
  filter(! Participant %in% c(8000,8012) & Timepoint !=5) %>% 
  mutate(Timepoint_label=ifelse(Timepoint %in% c(1,2),"Baseline","Maint"),
         Timepoint=as.factor(Timepoint)) %>% 
  gather(key=metab,value=metab_value,-X,-Tube_ID,-Sample_type,-Study,-Participant,-Timepoint,-Timepoint_label) %>%
  left_join(unique(select(samdf_diversity,Participant,Group))) %>% 
  select(Participant,Group,Timepoint,Timepoint_label,metab,metab_value)

metab_data_spread <- metab_data %>% 
  spread(key=metab,value=metab_value)

```


```{r}
paired_df_ppt <- function(df,tp_1,tp_2){
  df <- df %>% 
    arrange(Participant)
  temp_1 <- df %>% 
    filter(Timepoint_label==tp_1) 
  temp_2 <- df %>% 
    filter(Timepoint_label==tp_2)
  ppt_list <- intersect(temp_1$Participant,temp_2$Participant)
  
  df_paired <- bind_rows(
    filter(temp_1, Participant %in% ppt_list),
    filter(temp_2,Participant %in% ppt_list)
  )
  return(df_paired)
}

var_function <- function(df_spread,num_quantile){
  vars_vector <- df_spread %>% 
    select(-Participant,-Timepoint,-Group,-Timepoint_label) %>% 
    as.matrix() %>% 
    colVars()
  name_vector <- df_spread %>% 
    select(-Participant,-Timepoint,-Group,-Timepoint_label) %>% 
    colnames()
  filt_names <- name_vector[vars_vector>quantile(vars_vector)[num_quantile]] #top 75% varying proteins =2
  return(filt_names)
}
```

Paired siggenes end of intervention to baseline
-filtered to top 75% most varying metabolites (like the host proteins)
-no sig individual metabolites from baseline to end of intervention
```{r}
metab_data_fermented <- metab_data %>% 
  filter(Group=="Fermented") %>% 
  spread(key=metab,value=metab_value)

metab_fermented_names <- var_function(metab_data_fermented,2)
metab_data_fermented_filt <- metab_data %>% 
  filter(Group=="Fermented" & metab %in% metab_fermented_names) %>% 
  spread(key=metab,value=metab_value)

metab_data_fermented_paired <- paired_df_ppt(metab_data_fermented_filt,"Baseline","Maint")
set.seed(20)
siggenes_cl <- c((-1:-18), 1:18)
siggenes_data <- t(metab_data_fermented_paired[-c(1:4)])
siggenes_output <- siggenes::sam(data=siggenes_data,cl=siggenes_cl)
summary(siggenes_output)
delta_siggenes <- findDelta(siggenes_output, fdr = 0.05)

#fiber group
metab_data_fiber <- metab_data %>% 
  filter(Group=="Fiber") %>% 
  spread(key=metab,value=metab_value)

metab_fiber_names <- var_function(metab_data_fiber,2)
metab_data_fiber_filt <- metab_data %>% 
  filter(Group=="Fiber" & metab %in% metab_fiber_names) %>% 
  spread(key=metab,value=metab_value)

metab_data_fiber_paired <- paired_df_ppt(metab_data_fiber_filt,"Baseline","Maint")
set.seed(20)
siggenes_cl <- c((-1:-16), 1:16)
siggenes_data <- t(metab_data_fiber_paired[-c(1:4)])
siggenes_output <- siggenes::sam(data=siggenes_data,cl=siggenes_cl)
summary(siggenes_output)
delta_siggenes <- findDelta(siggenes_output, fdr = 0.05)
```

Beta diversity of metabolites from baseline to end of maint
-no sig. difference in clusters between timepoints
```{r}
metab_data_fermented_paired_raw <- metab_data_fermented_paired %>% 
  select(-c(Participant:Timepoint_label))
permanova <- adonis(metab_data_fermented_paired_raw ~ Timepoint_label, data = metab_data_fermented_paired, method='euclidean',na.rm = TRUE)
as.data.frame(permanova$aov.tab)["Timepoint_label", "Pr(>F)"]

metab_data_fiber_paired_raw <- metab_data_fiber_paired %>% 
  select(-c(Participant:Timepoint_label))
permanova <- adonis(metab_data_fiber_paired_raw ~ Timepoint_label, data = metab_data_fiber_paired, method='euclidean',na.rm = TRUE)
as.data.frame(permanova$aov.tab)["Timepoint_label", "Pr(>F)"]
```
PCA on metabolites
```{r}
pca_fermented <- prcomp(metab_data_fermented_paired_raw,scale=TRUE)
pca_fermented_df <- pca_fermented[["x"]] %>% as.data.frame %>% 
  cbind(select(metab_data_fermented_paired,Participant,Group,Timepoint_label),.)

ggplot(pca_fermented_df,aes(x=PC1,y=PC2,colour=Timepoint_label))+
  geom_point()

#plot single participant along the PC1 and PC2 axis
pca_fermented_df$Participant %<>% as.factor
ggplot(pca_fermented_df,aes(x=Participant,y=PC1,colour=Timepoint_label))+
  geom_point()+
  coord_flip()
ggplot(pca_fermented_df,aes(x=Participant,y=PC2,colour=Timepoint_label))+
  geom_point()+
  coord_flip()
```


Calculate metabolite difference across time for each participant
-PCA by group, no sig. difference in clustering (small trend)
```{r}
metab_data_diff <- metab_data %>% 
  # filter(Group %in% c("Fiber","Fermented")) %>% 
  # gather(key=metab,value=metab_value,-c(Participant:Timepoint_label)) %>% 
  select(-Timepoint) %>% 
  spread(key=Timepoint_label,value=metab_value) %>% 
  mutate(diff=Maint-Baseline)

var_function <- function(df_spread,num_quantile){
  vars_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    as.matrix() %>% 
    colVars()
  name_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    colnames()
  filt_names <- name_vector[vars_vector>quantile(vars_vector)[num_quantile]] #top 75% varying proteins =2
  return(filt_names)
}

metab_diff_names <- metab_data_diff %>% 
  select(-Baseline,-Maint) %>% 
  spread(key=metab,value=diff) %>%
  na.omit() %>% 
  var_function(.,2)

metab_data_filt_diff_spread <- metab_data_diff %>% 
  filter(metab %in% metab_diff_names) %>%
  select(-Baseline,-Maint) %>% 
  spread(key=metab,value=diff) %>% 
  na.omit()

pca_diff <- prcomp(select(metab_data_filt_diff_spread,-Participant,-Group),scale=TRUE)
pca_diff_df <- pca_diff[["x"]] %>% as.data.frame %>% 
  cbind(select(metab_data_filt_diff_spread,Participant,Group),.)

ggplot(pca_diff_df,aes(x=PC1,y=PC2,colour=Group))+
  geom_point()

pca_diff_df_raw <- pca_diff_df %>% 
  select(-c(Participant:Group)) 
permanova <- adonis(pca_diff_df_raw ~ Group, data = pca_diff_df, method='jaccard',na.rm = TRUE)
as.data.frame(permanova$aov.tab)["Group", "Pr(>F)"]
```

Correlate alpha diversity with each metabolite
```{r}
lme1by1HCW <- function(df,id, vectorNames, yVar,xColName){
  pValueList <- c()
  xList <- c()
  correValueList <- c()
  stdEList <-c()
  for (i in vectorNames){
    # df=df_raw
    # vectorNames=taxa_names
    # i=vectorNames[1]
    newdf <- data.frame(yVar = df[,yVar], xColName = df[, i], id = df[,id])
    #print(newdf)
    lmeData <- lme(yVar ~ xColName, data = newdf,random = ~1|id,na.action = na.omit)
    pVal = summary(lmeData)$tTable[2,5]
    corre = summary(lmeData)$tTable[2,1]
    stdE = summary(lmeData)$tTable[2,2]
    xList <- c(xList,i)
    pValueList <- c(pValueList,pVal)
    correValueList <- c(correValueList,corre)
    stdEList <- c(stdEList,stdE)
  }
  pTable <- c()
  pTable <- data.frame(xColName = xList, correlation = correValueList, pValue = pValueList, pValueAdj = p.adjust(pValueList,method="BH"),stdE=stdEList)
}


var_function <- function(df_spread,num_quantile){
  vars_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    as.matrix() %>% 
    colVars()
  name_vector <- df_spread %>% 
    select(-Participant,-Group) %>% 
    colnames()
  filt_names <- name_vector[vars_vector>quantile(vars_vector)[num_quantile]] #top 75% varying proteins =2
  return(filt_names)
}
```

```{r}
samdf_diversity_tp2 <- samdf_diversity %>% 
  filter(Timepoint==2) %>% 
  select(Participant,Group,Observed) %>% 
  dplyr::rename(Observed_2=Observed)

metab_data_alpha_div <- metab_data %>% 
  left_join(.,select(samdf_diversity,Participant,Group,Timepoint,Observed)) %>% 
  select(Participant,Group,Timepoint,Timepoint_label,Observed,metab,metab_value) %>% 
  left_join(.,samdf_diversity_tp2) %>% 
  mutate(Observed_corrected=ifelse(is.na(Observed) & Timepoint_label=="Baseline",Observed_2,Observed)) #if participant does not have an observed vlaue for tp=1, replace with tp2=2. Not doing this for missing maintenance tp=7

metab_data_alpha_div_fermented_filt <- metab_data_alpha_div %>% 
  filter(Group=="Fermented" & metab %in% metab_fermented_names) %>% 
  spread(key=metab,value=metab_value) %>% 
  na.omit()
metab_ferm_lme <- lme1by1HCW(metab_data_alpha_div_fermented_filt,"Participant",metab_fermented_names,"Observed_corrected",metab)

```
Correlate alpha diversity with PC space
```{r}
metab_data_alpha_div$Participant %<>% as.factor
pca_fermented_df_alpha_div <- inner_join(unique(select(metab_data_alpha_div,Participant,Group,Timepoint_label,Observed_corrected)),
                                         pca_fermented_df) 
lme_obj_pc1 <- lme(Observed_corrected ~ PC1, data = pca_fermented_df_alpha_div,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc1)$tTable
lme_obj_pc2 <- lme(Observed_corrected ~ PC2, data = pca_fermented_df_alpha_div,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc2)$tTable
lme_obj_pc3 <- lme(Observed_corrected ~ PC3, data = pca_fermented_df_alpha_div,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc3)$tTable

ggplot(pca_fermented_df_alpha_div,aes(x=PC2,y=Observed_corrected,colour=Timepoint_label))+
  geom_point()

pca_fermented_df_alpha_div_diff <- pca_fermented_df_alpha_div %>% 
  gather(key=Param, value=Value,-Participant,-Group,-Timepoint_label) %>% 
  spread(key=Timepoint_label,value=Value) %>% 
  mutate(Difference=Maint-Baseline) %>% 
  select(-Baseline,-Maint) %>% 
  spread(key=Param,value=Difference)

lme_obj_pc1 <- lme(Observed_corrected ~ PC1, data = pca_fermented_df_alpha_div_diff,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc1)$tTable
lme_obj_pc2 <- lme(Observed_corrected ~ PC2, data = pca_fermented_df_alpha_div_diff,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc2)$tTable
lme_obj_pc3 <- lme(Observed_corrected ~ PC3, data = pca_fermented_df_alpha_div_diff,random = ~1|Participant,na.action = na.omit)
summary(lme_obj_pc3)$tTable

ggplot(pca_fermented_df_alpha_div_diff,aes(x=PC2,y=Observed_corrected))+
  geom_point()

#plot loadings of pc2
pca_fermented_loadings <- pca_fermented[["rotation"]] %>% as.data.frame 

pc2_fermented_loadings <- pca_fermented_loadings %>% 
  mutate(metab=row.names(.),
         PC2_mag=abs(PC2)) %>% 
  select(metab,PC2,PC2_mag) %>% 
  arrange(-PC2)

ggplot(pc2_fermented_loadings[c(1:10),],aes(x=reorder(metab,PC2_mag),y=PC2_mag))+
  geom_bar(stat="identity")+
  coord_flip()

sum(pca_fermented_loadings$PC2^2)
```
Correlate metabolites with cytokines
```{r}
olink_data_gather <- olink_data %>% 
  gather(key=cytokine,value=cytokine_value,-Group,-Participant,-Timepoint) 
olink_data_fermented <- olink_data_gather %>% 
  filter(Group=="Fermented") %>% 
  mutate(cytokine=make.names(cytokine),
         Timepoint=as.factor(Timepoint)) %>% 
  spread(key=cytokine,value=cytokine_value) 


cytokine_list <- unique(olink_data_gather$cytokine) %>% make.names
cytokine_serum_metab_lme <- data.frame(cytokine=c(),
                                   xColName=c(),
                                   correlation=c(),
                                   pValue=c(),
                                   pValueAdj=c(),
                                   stdE=c())
for (i in cytokine_list){
  # i=cytokine_list[1]
  df_cytokine_metab <- inner_join(select(olink_data_fermented,Participant,Timepoint,all_of(i)),
                                  select(metab_data_fermented,-Timepoint_label)) %>% 
    na.omit() %>% 
    data.frame
  df_raw <- df_cytokine_metab %>% 
    select(-Participant,-Timepoint,-Group)
  df_raw <- df_raw[,colSums(df_raw) > 0]
  df_in <- df_cytokine_metab %>% select(Participant,colnames(df_raw))
  metab_names <- select(df_raw,-all_of(i)) %>% colnames
  metab_all_LME <- lme1by1HCW(df_in,"Participant",metab_names,i,metab) %>% mutate(cytokine=i)
  sig_metab <- metab_all_LME %>% filter(pValueAdj <= 0.05) 
  cytokine_serum_metab_lme <- bind_rows(cytokine_serum_metab_lme,sig_metab)
}

#calculate LME based on difference between baseline and maint for each participant
metab_data_fermented_diff <- metab_data_fermented %>% 
  gather(key=metab,value=metab_value,-Participant,-Group,-Timepoint,-Timepoint_label) %>% 
  select(-Timepoint) %>% 
  spread(key=Timepoint_label,value=metab_value) %>% 
  mutate(Diff=Maint-Baseline) %>% 
  select(-Baseline,-Maint) %>% 
  spread(key = metab, value=Diff)

olink_data_fermented_diff <- olink_data_fermented %>% 
  filter(Timepoint %in% c(1,6)) %>% 
  gather(key=cytokine,value=cytokine_value,-Participant,-Group,-Timepoint) %>% 
  spread(key=Timepoint,value=cytokine_value) %>% 
  mutate(Diff=`6`-`1`,
         cytokine=make.names(cytokine)) %>% 
  select(-`6`,-`1`) %>% 
  spread(key = cytokine, value=Diff)

cytokine_list <- select(olink_data_fermented_diff,-Participant,-Group) %>%colnames()
cytokine_serum_metab_diff_lme <- data.frame(cytokine=c(),
                                   xColName=c(),
                                   correlation=c(),
                                   pValue=c(),
                                   pValueAdj=c(),
                                   stdE=c())
for (i in cytokine_list){
  i=cytokine_list[1]
  df_cytokine_metab <- inner_join(select(olink_data_fermented_diff,Participant,all_of(i)),
                                  metab_data_fermented_diff) %>% 
    na.omit() %>% 
    data.frame
  df_raw <- df_cytokine_metab %>% 
    select(-Participant,-Group)
  df_raw <- df_raw[,colSums(df_raw) != 0]
  df_in <- df_cytokine_metab %>% select(Participant,colnames(df_raw))
  metab_names <- select(df_raw,-all_of(i)) %>% colnames
  metab_all_LME <- lme1by1HCW(df_in,"Participant",metab_names,i,metab) %>% mutate(cytokine=i)
  sig_metab <- metab_all_LME %>% filter(pValueAdj <= 0.05) 
  cytokine_serum_metab_diff_lme <- bind_rows(cytokine_serum_metab_diff_lme,sig_metab)
}
```

Correlate metabolites with microbial ASVs
-fermented group
```{r}
ASV_label_df <- df_ranks_gather %>% 
  select(ASV) %>% 
  unique() %>% 
  mutate(ASV_label=paste("ASV",c(1:dim(.)[1]),sep="_")) 
df_ranks_spread <- df_ranks_gather %>% 
  spread(key=ASV,value=rank_value)

df_ranks_fermented <- df_ranks_spread %>% 
  filter(Group=="Fermented")
taxa.raw <- select(df_ranks_fermented, -c(Participant:Timepoint))
filter.index1 <- apply(taxa.raw,2,function(X){sum(X>1)>0.25*length(X)}) #filter asvs to at least 10% of samples
taxa.filter <- taxa.raw[,filter.index1]
df_ranks_fermented <- select(df_ranks_fermented, c(Participant:Timepoint)) %>%
  bind_cols(., taxa.filter) %>% 
  mutate(Timepoint_value=as.integer(Timepoint))
asv_list <- colnames(taxa.filter)

ASV_serum_metab_lme_fermented <- data.frame(ASV_label=c(),
                                   xColName=c(),
                                   correlation=c(),
                                   pValue=c(),
                                   pValueAdj=c(),
                                   stdE=c())
for (i in asv_list){
  # i=ASV_list[1]
  print(i)
  df_metab <- inner_join(select(df_ranks_fermented,Participant,Timepoint,all_of(i)),
                                  select(metab_data_fermented,-Timepoint_label)) %>% 
    na.omit() %>% 
    data.frame
  df_raw <- df_metab %>% 
    select(-Participant,-Timepoint,-Group)
  df_raw <- df_raw[,colSums(df_raw) > 0]
  df_in <- df_metab %>% select(Participant,colnames(df_raw))
  metab_names <- select(df_raw,-all_of(i)) %>% colnames
  metab_all_LME <- lme1by1HCW(df_in,"Participant",metab_names,i,metab) %>% mutate(ASV=i)
  sig_metab <- metab_all_LME %>% filter(pValueAdj <= 0.05) 
  ASV_serum_metab_lme_fermented <- bind_rows(ASV_serum_metab_lme_fermented,sig_metab)
}

tax_table_df <- tax_table(ps_dada2_with_tree) %>% 
  as.data.frame() %>% 
  mutate(ASV=rownames(.))

ASV_serum_metab_lme_taxa_label_fermented <-ASV_serum_metab_lme_fermented %>% 
  left_join(.,tax_table_df)
# write.csv(ASV_serum_metab_lme_taxa_label_fermented,paste(save_path_figures,"ASV_serum_metab_lme_taxa_label_fermented.csv"))
ASV_serum_metab_lme_taxa_label_fermented <- read.csv(paste(save_path_figures,"ASV_serum_metab_lme_taxa_label_fermented.csv"))


df_ranks_fermented_gather <- df_ranks_fermented %>% 
  gather(key=ASV,value=rank_value,-c(Participant:Timepoint))
df_ranks_fermented_gather_filt <- df_ranks_fermented_gather %>% 
  filter(ASV %in% ASV_serum_metab_lme$cytokine) %>% 
  left_join(.,tax_table_df) %>% 
  mutate(taxa_name=paste(Family,Genus,Species,sep="_"))
```

```{r}
df_ranks_fiber <- df_ranks_spread %>% 
  filter(Group=="Fiber")
taxa.raw <- select(df_ranks_fiber, -c(Participant:Timepoint))
filter.index1 <- apply(taxa.raw,2,function(X){sum(X>1)>0.25*length(X)}) #filter asvs to at least 10% of samples
taxa.filter <- taxa.raw[,filter.index1]
df_ranks_fiber <- select(df_ranks_fiber, c(Participant:Timepoint)) %>%
  bind_cols(., taxa.filter) %>% 
  mutate(Timepoint_value=as.integer(Timepoint))
asv_list <- colnames(taxa.filter)

ASV_serum_metab_lme_fiber <- data.frame(ASV_label=c(),
                                   xColName=c(),
                                   correlation=c(),
                                   pValue=c(),
                                   pValueAdj=c(),
                                   stdE=c())
for (i in asv_list){
  # i=ASV_list[1]
  print(i)
  df_metab <- inner_join(select(df_ranks_fiber,Participant,Timepoint,all_of(i)),
                                  select(metab_data_fiber,-Timepoint_label)) %>% 
    na.omit() %>% 
    data.frame
  df_raw <- df_metab %>% 
    select(-Participant,-Timepoint,-Group)
  df_raw <- df_raw[,colSums(df_raw) > 0]
  df_in <- df_metab %>% select(Participant,colnames(df_raw))
  metab_names <- select(df_raw,-all_of(i)) %>% colnames
  metab_all_LME <- lme1by1HCW(df_in,"Participant",metab_names,i,metab) %>% mutate(ASV=i)
  sig_metab <- metab_all_LME %>% filter(pValueAdj <= 0.05) 
  ASV_serum_metab_lme_fiber <- bind_rows(ASV_serum_metab_lme_fiber,sig_metab)
}

ASV_serum_metab_lme_taxa_label_fiber <-ASV_serum_metab_lme_fiber %>% 
  left_join(.,tax_table_df)

#write.csv(ASV_serum_metab_lme_taxa_label_fiber,paste(save_path_figures,"ASV_serum_metab_lme_taxa_label_fiber.csv"))
ASV_serum_metab_lme_taxa_label_fiber <- read.csv(paste(save_path_figures,"ASV_serum_metab_lme_taxa_label_fiber.csv"))

```

Import taxa metab data from Han, Van Treuren et. al.
```{r}
# id_key <- read.delim("~/R/metabolomics_projects/strain_library_metab_info/metab_id_key_old_new.txt")
id_key_all <- read.delim("~/R/metabolomics_projects/strain_library_metab_info/metab_id_key_old_new_updated.txt")
id_key_compress <- id_key_all %>% 
  mutate(wvt_feature_new=ifelse(wvt_feature=="",feature_set,wvt_feature))

id_key <- id_key_compress %>% 
  select(hw_feature,wvt_feature_new)
genus_metab_lib <- read.delim("~/R/metabolomics_projects/strain_library_metab_info/Archive/genus_median_v01.txt")
family_metab_lib <- read.delim("~/R/metabolomics_projects/strain_library_metab_info/Archive/family_median_v01.txt")

genus_metab_lib_consume <- read.delim("~/R/metabolomics_projects/strain_library_metab_info/metabolitemodelscorrected/genus_median_v01.txt")

```

Find all of the metabolites that are produced by microbes at the genus level
```{r}
genus_metab_lib_gather <- genus_metab_lib %>% 
  gather(key=wvt_feature_new,value=med_value,-genus,-media) 

genus_metab_all <- genus_metab_lib_gather %>% 
  na.omit() %>% 
  left_join(.,id_key)

length(unique(genus_metab_all$hw_feature))


genus_metab_lib_consume_gather <- genus_metab_lib_consume %>% 
  gather(key=wvt_feature_new,value=med_value,-genus,-media) 

genus_metab_all_consume <- genus_metab_lib_consume_gather %>% 
  na.omit() %>% 
  left_join(.,id_key)

length(unique(genus_metab_all_consume$hw_feature))

```


Find intersection between 