---
title: "proteomics_parse_working"
author: "HCW"
date: "12/15/2019"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
library(magrittr)
library(matrixStats)
library(vegan)
```


Import data
```{r}
save_path_figures <- "../../plots/"
save_path_proteomics_raw <- "../../data/proteomics/raw/"
save_path_proteomics <- "../../data/proteomics/cleaned/"

proteomics_raw <- read.csv(paste(save_path_proteomics_raw,"proteomics_raw_All_reps_norm_no_Log.csv",sep="")) 
proteomics_mapping_file <- read.csv(paste(save_path_proteomics_raw,"pilot_stool_proteomics_sample_mappingfile.csv",sep="")) %>% 
  select(Participant,Group) %>% 
  unique()

proteomics_mapping_file$Participant %<>% as.character
proteomics_description <- proteomics_raw %>% 
  select(Accession,Description)
#bin the unique descriptions to description ids (collapse some parameters, ~1000 less)
proteomics_description_ids <- data.frame(Description=unique(proteomics_description$Description),Description_ID=paste("Desc_ID",c(1:length(unique(proteomics_description$Description))),sep="_"))

# saveRDS(proteomics_description,file = paste(save_path_proteomics, "proteo_description.rds", sep = ""))
# saveRDS(proteomics_description_ids,file = paste(save_path_proteomics, "proteo_description_ids.rds", sep = ""))

#sort proteins into host or microbe derived 
proteomics_host <- proteomics_description %>% 
  dplyr::filter(grepl("OS=Homo sapiens",Description)) %>% 
  left_join(.,proteomics_description_ids)
proteomics_microbe <- proteomics_description %>% 
  dplyr::filter(!(grepl("OS=Homo sapiens",Description)))%>% 
  left_join(.,proteomics_description_ids)

#average two runs, take the log2+1 of the average
proteomics_data <- proteomics_raw %>%
  gather(key=SampleID_temp,value=proteo_value,-Accession,-Description) %>%
  separate(col=SampleID_temp,into=c("SampleID","Run","Timepoint_num"),sep="_",remove=FALSE) %>%
  mutate(Participant=substring(SampleID, 2)) %>%
  select(Participant,Run,Timepoint_num,Accession,Description,proteo_value) %>%
  spread(key=Run,value=proteo_value) %>%
  dplyr::rename(Run_1=`1`,Run_2=`2`) %>%
  mutate(avg_log=log2((Run_1+Run_2)/2+1)) %>%
  select(-Run_1,-Run_2) %>%
  mutate(Timepoint=recode(Timepoint_num,"1"=1,"2"=2,"3"=3,"4"=6,"5"=7)) %>%
  right_join(proteomics_mapping_file,.) %>%
  select(-Timepoint_num) %>%
  filter(!(Participant %in% c("8000","8012")))
# saveRDS(proteomics_data,paste(save_path_proteomics,"proteomics_data.rds"))

#summarize the proteins by summing all that have the same description
proteomics_data_sum_desc <- proteomics_data %>% 
  group_by(Participant,Group,Timepoint,Description) %>% 
  dplyr::summarize(avg_log_sum=sum(avg_log)) %>% 
  right_join(.,proteomics_description_ids)

#filter to host and microbe proteins
proteo_host_gather <- proteomics_data %>% 
  filter(Accession %in% proteomics_host$Accession)
#length(proteomics_host$Accession) == length(unique(proteomics_host$Description)) 
#all host proteins have unique descriptions, don't need to filter to host sum desc

proteo_microbe_gather <- proteomics_data %>% 
  filter(Accession %in% proteomics_microbe$Accession)


#filter sum descriptions for microbe data to collapse number of parameters (decrease by ~1000)
proteo_microbe_sum_desc_gather <- proteomics_data_sum_desc %>% 
  filter(Description %in% unique(proteomics_microbe$Description))


#futher normalize the samples as a percent of total protein in each sample (participant, timepoint)
proteo_host_gather_norm <- proteo_host_gather %>% 
  group_by(Participant,Group,Timepoint) %>% 
  dplyr::summarize(sample_protein_sum=sum(avg_log)) %>% 
  left_join(proteo_host_gather,.) %>% 
  mutate(avg_log_norm=avg_log/sample_protein_sum)
proteo_host_norm_spread <- proteo_host_gather_norm %>% 
  select(-Description,-avg_log,-sample_protein_sum) %>% 
  spread(key=Accession,value=avg_log_norm)
#saveRDS(proteo_host_gather_norm,file = paste(save_path_proteomics, "proteo_host_gather_norm.rds", sep = ""))

proteo_microbe_sum_desc_gather_norm <- proteo_microbe_sum_desc_gather %>% 
  group_by(Participant,Group,Timepoint) %>% 
  dplyr::summarize(sample_protein_sum=sum(avg_log_sum)) %>% 
  left_join(proteo_microbe_sum_desc_gather,.) %>% 
  mutate(avg_log_norm=avg_log_sum/sample_protein_sum)
proteo_microbe_sum_desc_norm_spread <- proteo_microbe_sum_desc_gather_norm %>% 
  select(-Description,-avg_log_sum,-sample_protein_sum) %>% 
  spread(key=Description_ID,value=avg_log_norm)

#saveRDS(proteo_host_gather_norm,file = paste(save_path_proteomics, "proteo_host_gather_norm.rds", sep = ""))
#saveRDS(proteo_microbe_sum_desc_gather_norm,file = paste(save_path_proteomics, "proteo_microbe_sum_desc_gather_norm.rds", sep = ""))
```

Calculate all variables for each participant as a difference from timepoint 1
```{r}
#calculate difference from timepoint 1
diff_from_1 <- function(spread_df,key_name,value_name,base_value_name,new_name){
  df_gather <- spread_df %>% 
    gather(key=key_name,value=value_name,-Participant,-Timepoint,-Group) %>% 
    arrange(Participant)
  df_gather_1 <- df_gather %>% 
    filter(Timepoint==1) %>% 
    dplyr::rename(base_value_name=value_name) %>% 
    select(-Timepoint)
  df_gather_not_1 <- df_gather %>% 
    filter(Timepoint!=1)
  df_gather_diff <- left_join(df_gather_1,df_gather_not_1) %>% 
    mutate(new_name=value_name-base_value_name) %>% 
    select(-base_value_name,-value_name) 
  df_spread_diff <- df_gather_diff %>% 
    spread(key=key_name,value=new_name)
  return(df_spread_diff)
}

proteo_host_norm_diff <- diff_from_1(proteo_host_norm_spread,key_name=Accession,value_name=avg_log,base_value_name=avg_log_norm_1,new_name=avg_log_norm_diff)
proteo_host_diff_norm_gather <- proteo_host_norm_diff %>% 
  gather(key=Accession,value=avg_log_norm,-Participant,-Group,-Timepoint)

proteo_host_norm_diff_fiber <- proteo_host_norm_diff %>% 
  filter(Group=="Fiber")
proteo_host_norm_diff_fermented <- proteo_host_norm_diff %>% 
  filter(Group=="Fermented")

proteo_microbe_desc_norm_diff <- diff_from_1(ungroup(proteo_microbe_sum_desc_norm_spread),key_name=Description_ID,value_name=avg_log,base_value_name=avg_log_norm_1,new_name=avg_log_sum_norm_diff)
proteo_microbe_desc_diff_norm_gather <- proteo_microbe_desc_norm_diff %>% 
  gather(key=Description_ID,value=avg_log_norm_sum,-Participant,-Group,-Timepoint)

proteo_microbe_desc_norm_diff_fiber <- proteo_microbe_desc_norm_diff %>% 
  filter(Group=="Fiber")
proteo_microbe_desc_norm_diff_fermented <- proteo_microbe_desc_norm_diff %>% 
  filter(Group=="Fermented")
```

Subset parameters to most varying, host and microbe
  - most variable in both fiber and fermented group
  - most variable in fiber group only 
  - most variable in fermented group only
```{r}
var_function <- function(df_spread,num_quantile){
  vars_vector <- df_spread %>% 
    select(-Participant,-Timepoint,-Group) %>% 
    as.matrix() %>% 
    colVars()
  name_vector <- df_spread %>% 
    select(-Participant,-Timepoint,-Group) %>% 
    colnames()
  filt_names <- name_vector[vars_vector>quantile(vars_vector)[num_quantile]] #top 75% varying proteins =2
  return(filt_names)
}


proteo_host_diff_names <- var_function(proteo_host_norm_diff,2)
proteo_host_diff_names_fiber <- var_function(proteo_host_norm_diff_fiber,2)
proteo_host_diff_names_fermented <- var_function(proteo_host_norm_diff_fermented,2)

proteo_microbe_desc_diff_names <- var_function(proteo_microbe_desc_norm_diff,3)
proteo_microbe_desc_diff_names_fiber <- var_function(proteo_microbe_desc_norm_diff_fiber,3)
proteo_microbe_desc_diff_names_fermented <- var_function(proteo_microbe_desc_norm_diff_fermented,3)

#filter the differences from timepoint 1 data frames 
proteo_host_diff_filt <- proteo_host_diff_norm_gather %>% 
  filter(Accession %in% proteo_host_diff_names) %>% 
  spread(key=Accession, value=avg_log_norm)
proteo_host_diff_filt_fiber <- proteo_host_diff_norm_gather %>% 
  filter(Group=="Fiber" & Accession %in% proteo_host_diff_names_fiber) %>% 
  spread(key=Accession, value=avg_log_norm)
proteo_host_diff_filt_fermented <- proteo_host_diff_norm_gather %>% 
  filter(Group=="Fermented" & Accession %in% proteo_host_diff_names_fermented) %>% 
  spread(key=Accession, value=avg_log_norm)

proteo_microbe_desc_diff_filt <- proteo_microbe_desc_diff_norm_gather %>% 
  filter(Description_ID %in% proteo_microbe_desc_diff_names) %>% 
  spread(key=Description_ID, value=avg_log_norm_sum)
proteo_microbe_desc_diff_filt_fiber <- proteo_microbe_desc_diff_norm_gather %>% 
  filter(Group=="Fiber" & Description_ID %in% proteo_microbe_desc_diff_names) %>% 
  spread(key=Description_ID, value=avg_log_norm_sum)
proteo_microbe_desc_diff_filt_fermented <- proteo_microbe_desc_diff_norm_gather %>% 
  filter(Group=="Fermented" & Description_ID %in% proteo_microbe_desc_diff_names) %>% 
  spread(key=Description_ID, value=avg_log_norm_sum)

#filter the dataframes that includes timepoint 1 (not differences)
proteo_host_filt <- proteo_host_gather_norm %>% 
  filter(Accession %in% proteo_host_diff_names) %>% 
  select(-Description,-sample_protein_sum,-avg_log) %>% 
  spread(key=Accession, value=avg_log_norm)
proteo_host_filt_fiber <- proteo_host_gather_norm %>% 
  filter(Group=="Fiber" & Accession %in% proteo_host_diff_names) %>% 
  select(-Description,-sample_protein_sum,-avg_log) %>% 
  spread(key=Accession, value=avg_log_norm)
proteo_host_filt_fermented <- proteo_host_gather_norm %>% 
  filter(Group=="Fermented" & Accession %in% proteo_host_diff_names) %>% 
  select(-Description,-sample_protein_sum,-avg_log) %>% 
  spread(key=Accession, value=avg_log_norm)

proteo_microbe_desc_filt <- proteo_microbe_sum_desc_gather_norm %>% 
  filter(Description_ID %in% proteo_microbe_desc_diff_names) %>% 
  select(-Description,-avg_log_sum,-sample_protein_sum) %>% 
  spread(key=Description_ID, value=avg_log_norm)
proteo_microbe_desc_filt_fiber <- proteo_microbe_sum_desc_gather_norm %>% 
  filter(Group=="Fiber" & Description_ID %in% proteo_microbe_desc_diff_names) %>% 
  select(-Description,-avg_log_sum,-sample_protein_sum) %>% 
  spread(key=Description_ID, value=avg_log_norm)
proteo_microbe_desc_filt_fermented <- proteo_microbe_sum_desc_gather_norm %>% 
  filter(Group=="Fermented" & Description_ID %in% proteo_microbe_desc_diff_names) %>% 
  select(-Description,-avg_log_sum,-sample_protein_sum) %>% 
  spread(key=Description_ID, value=avg_log_norm)


# saveRDS(proteo_host_diff_filt,file = paste(save_path_proteomics, "proteo_host_diff_filt.rds", sep = ""))
# saveRDS(proteo_host_diff_filt_fiber,file = paste(save_path_proteomics, "proteo_host_diff_filt_fiber.rds", sep = ""))
# saveRDS(proteo_host_diff_filt_fermented,file = paste(save_path_proteomics, "proteo_host_diff_filt_fermented.rds", sep = ""))
# saveRDS(proteo_microbe_desc_diff_filt,file = paste(save_path_proteomics, "proteo_microbe_desc_diff_filt.rds", sep = ""))
# saveRDS(proteo_microbe_desc_diff_filt_fiber,file = paste(save_path_proteomics, "proteo_microbe_desc_diff_filt_fiber.rds", sep = ""))
# saveRDS(proteo_microbe_desc_diff_filt_fermented,file = paste(save_path_proteomics, "proteo_microbe_desc_diff_filt_fermented.rds", sep = ""))
# saveRDS(proteo_host_filt,file = paste(save_path_proteomics, "proteo_host_filt.rds", sep = ""))
# saveRDS(proteo_host_filt_fiber,file = paste(save_path_proteomics, "proteo_host_filt_fiber.rds", sep = ""))
# saveRDS(proteo_host_filt_fermented,file = paste(save_path_proteomics, "proteo_host_filt_fermented.rds", sep = ""))
# saveRDS(proteo_microbe_desc_filt,file = paste(save_path_proteomics, "proteo_microbe_desc_filt.rds.rds", sep = ""))
# saveRDS(proteo_microbe_desc_filt_fiber,file = paste(save_path_proteomics, "proteo_microbe_desc_filt_fiber.rds", sep = ""))
# saveRDS(proteo_microbe_desc_filt_fermented,file = paste(save_path_proteomics, "proteo_microbe_desc_filt_fermented.rds", sep = ""))

```

Prepare host proteo data for IPA analysis 
  - ID column
  - log2foldchange column (average for a group)
  - p-value for differential analysis 
```{r}
proteo_host_norm_gather <- proteo_host_norm_spread %>% 
  gather(key=Accession,value=log_ratio,-Participant,-Group,-Timepoint) %>% 
  filter(Group=="Fermented")
# proteo_host_norm_spread <- proteo_host_norm_gather %>% 
#   select(-Timepoint) %>% 
#   spread(key=Accession,value=log_ratio)
proteo_host_accession_list <- unique(proteo_host_norm_gather$Accession)
tp_list <- unique(filter(proteo_host_norm_gather,Timepoint!=1)$Timepoint)
#calculate pvalues 
proteo_host_paired_df <- data.frame(Accession=c(),
                                    Timepoint=c(),
                                    pvalue=c())
for(j in tp_list){
  # i=proteo_host_accession_list[1]
  pvalue_list=c()
  for(i in proteo_host_accession_list){
    # j=2
    df_1=filter(proteo_host_norm_gather,Timepoint==1 & Accession==i)
    df_j=filter(proteo_host_norm_gather,Timepoint==j & Accession==i)
    ppt_list=intersect(df_1$Participant,df_j$Participant)
    df_1=df_1 %>% 
      filter(Participant %in% ppt_list) %>% 
      arrange(Participant)
    df_j=df_j %>% 
      filter(Participant %in% ppt_list) %>% 
      arrange(Participant)
    vect_1=df_1$log_ratio
    vect_j=df_j$log_ratio
    pvalue=t.test(vect_1,vect_j,paired=TRUE)$p.value
    pvalue_list=c(pvalue_list,pvalue)
  }
  temp_df <- data.frame(Accession=proteo_host_accession_list,
                        Timepoint=j,
                        pvalue=pvalue_list)
  proteo_host_paired_df <- bind_rows(proteo_host_paired_df,temp_df)
}

proteo_host_paired_df_all <- proteo_host_paired_df %>% 
  mutate(Timepoint_label=paste("pvalue_Timepoint",Timepoint,sep="_")) %>% 
  select(-Timepoint) %>% 
  spread(key=Timepoint_label,value=pvalue)
  
proteo_host_diff_norm_gather_group_all <- proteo_host_diff_norm_gather %>% 
  filter(Group=="Fermented") %>% 
  group_by(Accession,Timepoint) %>% 
  dplyr::summarize(avg_log_norm_all=mean(avg_log_norm)) %>% 
  mutate(Timepoint_label=paste("log2_Timepoint",Timepoint,sep="_")) %>% 
  select(-Timepoint) %>% 
  spread(key=Timepoint_label,value=avg_log_norm_all)

proteo_host_paired_df_all <- full_join(proteo_host_diff_norm_gather_group_all,proteo_host_paired_df_all)
write.csv(proteo_host_paired_df_all,"~/R/proteo_host_paired_df_fermented.csv")
```

```{r}
proteo_host_go_df <- read.delim("~/R/pantherGeneList (1).txt",header=FALSE)
colnames(proteo_host_go_df) <- c("Gene_ID","Accession","Gene_name","Family_subfamily","Protein_class","Molecular_function","Biological_process","Pathway")

protein_class_list <- unique(unlist(strsplit(as.character(unique(proteo_host_go_df$Protein_class)),split=";"))) #length=58
molecular_functions_list <- unique(unlist(strsplit(as.character(unique(proteo_host_go_df$Molecular_function)),split=";"))) #length=93
biological_process_list <- unique(unlist(strsplit(as.character(unique(proteo_host_go_df$Biological_process)),split=";"))) #length=156
pathway_list <- unique(unlist(strsplit(as.character(unique(proteo_host_go_df$Pathway)),split=";"))) #length=47

nested_list_df_function <- function(vector,split_char,naming_vector){
  # vector=proteo_host_go_df$Molecular_function
  # split_char=";"
  # naming_vector=proteo_host_go_df$Accession
  working_list <- strsplit(as.character(vector),split=split_char)
  names(working_list) <- naming_vector
  flatten_list <- ldply(working_list,data.frame,.id="Name")
  colnames(flatten_list) <- c("Analyte","Description")
  flatten_list_all <- left_join(data.frame(Analyte=unique(proteo_host_gather$Accession)),flatten_list)
  list_df <- dcast(flatten_list,formula = Analyte ~ Description, fun.aggregate = length)
  # rownames(list_df) <- list_df$Analyte
  # list_df <- select(list_df,-Analyte)
  return(list_df)
}

protein_class_host_proteo_df <- nested_list_df_function(proteo_host_go_df$Protein_class,
                                                              ";",
                                                              proteo_host_go_df$Accession)
molecular_functions_host_proteo_df <- nested_list_df_function(proteo_host_go_df$Molecular_function,
                                                              ";",
                                                              proteo_host_go_df$Accession)
biological_processes_host_proteo_df <- nested_list_df_function(proteo_host_go_df$Biological_process,
                                                              ";",
                                                              proteo_host_go_df$Accession)
pathway_list_host_proteo_df <- nested_list_df_function(proteo_host_go_df$Pathway,
                                                              ";",
                                                              proteo_host_go_df$Accession)
# saveRDS(protein_class_host_proteo_df,"~/R/fiber-fermented-study/data/proteomics/cleaned/host_proteomics_annotations/protein_class_host_proteo_df.rds")
# saveRDS(molecular_functions_host_proteo_df,"~/R/fiber-fermented-study/data/proteomics/cleaned/host_proteomics_annotations/molecular_functions_host_proteo_df.rds")
# saveRDS(biological_processes_host_proteo_df,"~/R/fiber-fermented-study/data/proteomics/cleaned/host_proteomics_annotations/biological_processes_host_proteo_df.rds")
# saveRDS(pathway_list_host_proteo_df,"~/R/fiber-fermented-study/data/proteomics/cleaned/host_proteomics_annotations/pathway_list_host_proteo_df.rds")

# saveRDS(proteo_host_go_df,"~/R/fiber-fermented-study/data/proteomics/cleaned/host_proteomics_annotations/proteo_host_go_df.rds")

```

