---
title: "O-link analysis and plotting"
author: "GK Fragiadakis"
date: "8/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

O-link analysis that's included in publication. Raw data, processed data and data processing script can also be found in this repo. 

### Significance analysis of O-link data

```{r}
library(tidyverse)
library(siggenes)
library(knitr)
library(RColorBrewer)

data_directory <- "../../data/Olink/cleaned/"
plot_directory <- "../../plots/"

olink_data <- read_csv(paste(data_directory, "olink_data_cleaned.csv", sep = "")) %>% mutate(Timepoint = as.numeric(Timepoint))

```

Running SAM from siggenes library on fermented food cohort comparing baseline to end of maintenance:

```{r}

select_participants <- function(data, time1, time2){
  
   ppts_time1 <- data %>% dplyr::filter(Timepoint == time1) %>% select(Participant) %>% arrange(Participant) %>% distinct()
   ppts_time2 <- data %>% dplyr::filter(Timepoint == time2) %>% select(Participant) %>% arrange(Participant) %>% distinct()
   
   selected_participants <- intersect(ppts_time1, ppts_time2)

   return(selected_participants$Participant)
}

prep_sam <- function(data, selected_participants, time1, time2, metadata_columns){
  
  
  analysis_data <- data %>% 
              dplyr::filter(Participant %in% selected_participants) %>% 
              dplyr::filter(Timepoint == time1 | Timepoint == time2)
  
  sam_key <- analysis_data %>% 
                mutate(Sam_index = Timepoint) %>% 
                select(Participant, Timepoint, Sam_index)
  
  sam_key[sam_key$Timepoint == time1, "Sam_index"] <- -1
  
  sam_key[sam_key$Timepoint == time2, "Sam_index"] <- 1
  
  k = 1
  
  for (i in unique(sam_key$Participant)){
    
    sam_key[sam_key$Participant == i, "Participant_ID"] <- k
    
    k <- k + 1
  }
  
  sam_key <- sam_key %>% mutate(SAM_ID = Sam_index*Participant_ID) %>% select(Participant, Timepoint, SAM_ID)
  
  analysis_data <- left_join(sam_key, analysis_data)
  
  x <- analysis_data %>% select_(.dots = paste("-", metadata_columns)) %>% select(-SAM_ID) %>% t()
  y <- analysis_data$SAM_ID
    
    return(list(x = x,y = y))
  
}

run_sam <- function(full_data, diet, time1, time2, metadata_columns){
  
  data <- full_data %>% dplyr::filter(Group == diet)
  
  selected_participants <- select_participants(data, time1, time2)
  
  input_list <- prep_sam(data, selected_participants, time1, time2, metadata_columns)
  
  x <- input_list[["x"]]
  
  y <- input_list[["y"]]
  
  siggenes_model <- siggenes::sam(data = x, cl = y, gene.names = rownames(x), rand = 123)
  
  return(list(model = siggenes_model, x = x, y = y, selected_participants = selected_participants))
  
}


fermented_info <- run_sam(full_data = olink_data, diet = "Fermented", time1 = 1, time2 = 6,
                      metadata_columns = colnames(select(olink_data, Participant, Timepoint, Group))) 

n_fermented <- length(fermented_info[["selected_participants"]])

findDelta(fermented_info[["model"]] , fdr = 0.05)

# input delta by looking manually

selected_fermented_model <- summary(fermented_info[["model"]], 1.355231)

# Features with q-value < .05

fermented_signaling_significant <- selected_fermented_model@mat.sig %>% 
                                          mutate(Feature = rownames(selected_fermented_model@mat.sig)) %>% 
                                          dplyr::filter(`q.value` < .05) %>% 
                                          select(Feature, `d.value`, `q.value`)

kable(fermented_signaling_significant, caption = paste("CyTOF signaling features significant by q-value, fermented food cohort, n = ", n_fermented, sep = ""))


```


Plotting the results from the fermented group:

```{r, fig.width = 9}

full_fermented_data <- olink_data %>% 
  dplyr::filter(Group == "Fermented") %>% 
  select(-Group) %>%
  dplyr::filter(Participant %in% fermented_info$selected_participants) %>%
  gather(key = Analyte, value = value, -c(Participant:Timepoint))

graph_list <- list()

for (analyte in fermented_signaling_significant$Feature){
  
analyte_df <- full_fermented_data %>% 
              dplyr::filter(Analyte == analyte) %>% 
              dplyr::select(-Analyte) %>% 
              spread(., key = Timepoint, value = value)

fold_df <- analyte_df %>%  
            dplyr::mutate(`-3` = `1`-`1`,
                           `0` = `2` - `1`, 
                          `6_` = `4`- `1`, 
                          `8` = `5`-`1`, 
                         `10` = `6`-`1`) %>%
            dplyr::select(-c(`1`:`6`)) %>%
            dplyr::rename()
            gather(., key = "Time", value = "difference", `-3wks`:`10wks`) %>%
            mutate(Time = factor(Time, levels = c("-3wks", "0wks", "6wks", "8wks", "10wks")))

p <- ggplot(fold_df, aes(x=Time, y=difference)) + geom_boxplot(colour = "#bebada", fill = "#bebada", alpha = 0.5) + 
          ggtitle(analyte) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(), axis.line = element_line(colour = "black"))  +
          theme(plot.title = element_text(hjust = 0.5), axis.title.x=element_blank())

graph_list[[analyte]] <- p

}

library(ggpubr)

ggarrange(graph_list[[1]], graph_list[[2]], graph_list[[3]], graph_list[[4]], graph_list[[5]],
          graph_list[[6]], graph_list[[7]], graph_list[[8]], graph_list[[9]], graph_list[[10]],
          graph_list[[11]], ncol = 4, nrow = 3)

ggsave(paste(plot_directory, "olink_significant_cytokines_over_time.pdf", sep = ""), useDingbats = FALSE, height = 5, width = 9)

```

Running SAM from siggenes library on fiber food cohort, comparing baseline to end of maintenance:

```{r}

fiber_info <- run_sam(full_data = olink_data, diet = "Fiber", time1 = 1, time2 = 6,
                      metadata_columns = colnames(select(olink_data, Participant, Timepoint, Group))) 


findDelta(fiber_info[["model"]] , fdr = 0.05)

n_fiber <- length(fiber_info[["selected_participants"]])

```

Nothing significantly changed in fiber group (n = `r as.character(n_fiber)`). 


