---
title: "O-link analysis and plotting"
author: "GK Fragiadakis"
date: "8/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

O-link analysis that's included in publication. Raw data, processed data and data processing script can also be found in this repo. 

### Significance analysis of O-link data

```{r}
library(tidyverse)
library(siggenes)
library(knitr)
library(RColorBrewer)

data_directory <- "../../data/Olink/cleaned/"
plot_directory <- "../../plots/"

olink_data <- read_csv(paste(data_directory, "olink_data_cleaned.csv", sep = "")) %>% mutate(Timepoint = as.numeric(Timepoint))

```

Running SAM from siggenes library on fermented food cohort comparing baseline to end of maintenance:

```{r}

select_participants <- function(data, time1, time2){
  
   ppts_time1 <- data %>% dplyr::filter(Timepoint == time1) %>% select(Participant) %>% arrange(Participant) %>% distinct()
   ppts_time2 <- data %>% dplyr::filter(Timepoint == time2) %>% select(Participant) %>% arrange(Participant) %>% distinct()
   
   selected_participants <- intersect(ppts_time1, ppts_time2)

   return(selected_participants$Participant)
}

prep_sam <- function(data, selected_participants, time1, time2, metadata_columns){
  
  
  analysis_data <- data %>% 
              dplyr::filter(Participant %in% selected_participants) %>% 
              dplyr::filter(Timepoint == time1 | Timepoint == time2)
  
  sam_key <- analysis_data %>% 
                mutate(Sam_index = Timepoint) %>% 
                select(Participant, Timepoint, Sam_index)
  
  sam_key[sam_key$Timepoint == time1, "Sam_index"] <- -1
  
  sam_key[sam_key$Timepoint == time2, "Sam_index"] <- 1
  
  k = 1
  
  for (i in unique(sam_key$Participant)){
    
    sam_key[sam_key$Participant == i, "Participant_ID"] <- k
    
    k <- k + 1
  }
  
  sam_key <- sam_key %>% mutate(SAM_ID = Sam_index*Participant_ID) %>% select(Participant, Timepoint, SAM_ID)
  
  analysis_data <- left_join(sam_key, analysis_data)
  
  x <- analysis_data %>% select_(.dots = paste("-", metadata_columns)) %>% select(-SAM_ID) %>% t()
  y <- analysis_data$SAM_ID
    
    return(list(x = x,y = y))
  
}

run_sam <- function(full_data, diet, time1, time2, metadata_columns){
  
  data <- full_data %>% dplyr::filter(Group == diet)
  
  selected_participants <- select_participants(data, time1, time2)
  
  input_list <- prep_sam(data, selected_participants, time1, time2, metadata_columns)
  
  x <- input_list[["x"]]
  
  y <- input_list[["y"]]
  
  siggenes_model <- siggenes::sam(data = x, cl = y, gene.names = rownames(x), rand = 123)
  
  return(list(model = siggenes_model, x = x, y = y, selected_participants = selected_participants))
  
}


fermented_info <- run_sam(full_data = olink_data, diet = "Fermented", time1 = 1, time2 = 6,
                      metadata_columns = colnames(select(olink_data, Participant, Timepoint, Group))) 

n_fermented <- length(fermented_info[["selected_participants"]])

findDelta(fermented_info[["model"]] , fdr = 0.05)
findDelta(fermented_info[["model"]] , fdr = 0.10)


# input delta by looking manually

selected_fermented_model <- summary(fermented_info[["model"]], 1.355231)
selected_fermented_model <- summary(fermented_info[["model"]], 1.251708)


# Features with q-value < .1

fermented_significant <- selected_fermented_model@mat.sig %>% 
                                          mutate(Feature = rownames(selected_fermented_model@mat.sig)) %>% 
                                          dplyr::filter(`q.value` < .1) %>% 
                                          select(Feature, `d.value`, `q.value`)

kable(fermented_significant, caption = paste("Olink signaling features significant by q-value, fermented food cohort, n = ", n_fermented, sep = ""))


```


Plotting the results from the fermented group:

```{r, fig.width = 9}

full_fermented_data <- olink_data %>% 
  dplyr::filter(Group == "Fermented") %>% 
  select(-Group) %>%
  dplyr::filter(Participant %in% fermented_info$selected_participants) %>%
  gather(key = Analyte, value = value, -c(Participant:Timepoint))

graph_list <- list()

for (analyte in fermented_significant$Feature){
  
analyte_df <- full_fermented_data %>% 
              dplyr::filter(Analyte == analyte) %>% 
              dplyr::select(-Analyte) %>% 
              spread(., key = Timepoint, value = value)

fold_df <- analyte_df %>%  
            dplyr::mutate(`-3` = `1`-`1`,
                           `0` = `2` - `1`, 
                          `6_` = `4`- `1`, 
                          `8` = `5`-`1`, 
                         `10` = `6`-`1`) %>%
            dplyr::select(-c(`1`:`6`)) %>%
            dplyr::rename(`6` = `6_`) %>%
            gather(., key = "Time", value = "difference", `-3`:`10`) %>%
            mutate(Time = factor(Time, levels = c("-3", "0", "6", "8", "10"))) %>%
            rename(Weeks = Time)

p <- ggplot(fold_df, aes(x=Weeks, y=difference)) + geom_boxplot(colour = "#bebada", fill = "#bebada", alpha = 0.5) + 
          ggtitle(analyte) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.background = element_blank(), axis.line = element_line(colour = "black"))  +
          theme(plot.title = element_text(hjust = 0.5))

graph_list[[analyte]] <- p

}

library(ggpubr)

args_list <- graph_list
args_list[["ncol"]] <- 5
args_list[["nrow"]] <- 4
do.call(ggarrange, args_list)

ggsave(paste(plot_directory, "olink_significant_cytokines_over_time_separate_plots.pdf", sep = ""), useDingbats = FALSE, height = 5, width = 9)

```

Additional plotting for figure: 

```{r}

sigg_fermented_data <- full_fermented_data %>% 
                        filter(Analyte %in% fermented_significant$Feature)

sigg_fermented_data$Timepoint %<>% as.factor()

sigg_fermented_data_diff <- sigg_fermented_data %>% 
                                    spread(key=Timepoint,value=value) %>% 
                                    dplyr::mutate(`-3` = `1`-`1`,
                                                  `0` = `2` - `1`, 
                                                  `6_` = `4`- `1`, 
                                                  `8` = `5`-`1`, 
                                                  `10` = `6`-`1`) %>%
                                    dplyr::select(-c(`1`:`6`)) %>%
                                    dplyr::rename(`6` = `6_`) %>%
                                    gather(., key = "Time", value = "difference", `-3`:`10`) %>%
                                    mutate(Time = factor(Time, levels = c("-3", "0", "6", "8", "10"))) %>%
                                    rename(Weeks = Time) 
                                    # filter(Analyte != "FGF-21",Weeks!="-3")

sigg_fermented_data_diff$Weeks %<>% as.factor

sigg_fermented_data_diff_no_FGF <- sigg_fermented_data_diff %>% filter(Analyte != "FGF-21",Weeks!="-3")
sigg_fermented_data_diff_FGF <- sigg_fermented_data_diff %>% filter(Analyte == "FGF-21",Weeks!="-3")


p <- ggplot(sigg_fermented_data_diff_no_FGF,aes(x=Analyte,y=difference,fill=Weeks))+
  geom_boxplot(color="#525252") +
  theme_classic()+ 
  theme(text = element_text(size=20),axis.text.x = element_text(angle=45, hjust=1)) +
  scale_fill_manual(values = c("#fcfbfd","#fcfbfd","#efedf5","#dadaeb","#bebada"))+
  ylab("Cytokine Difference from Week -3")+
  xlab("Inflammatory Cytokine") 

plot(p)
ggsave(paste(plot_directory, "olink_significant_cytokines_over_time_single_plot.pdf", sep = ""),width = 18,height=9.2,useDingbats=FALSE)


p <- ggplot(sigg_fermented_data_diff_FGF,aes(x=Analyte,y=difference,fill=Weeks))+
  geom_boxplot(color="#525252") +
  theme_classic()+ 
  theme(text = element_text(size=20),axis.text.x = element_text(angle=45, hjust=1)) +
  scale_fill_manual(values = c("#fcfbfd","#fcfbfd","#efedf5","#dadaeb","#bebada"))+
  ylab("Cytokine Difference from Week -3")+
  xlab("Inflammatory Cytokine") 

plot(p)
ggsave(paste(plot_directory, "olink_significant_cytokines_over_time_FGF.pdf", sep = ""),width = 3,height=9.2,useDingbats=FALSE)

```



Running SAM from siggenes library on fiber food cohort, comparing baseline to end of maintenance:

```{r}

fiber_info <- run_sam(full_data = olink_data, diet = "Fiber", time1 = 1, time2 = 6,
                      metadata_columns = colnames(select(olink_data, Participant, Timepoint, Group))) 


findDelta(fiber_info[["model"]] , fdr = 0.1)
selected_fiber_model <- summary(fiber_info[["model"]], 0.512142 )

n_fiber <- length(fiber_info[["selected_participants"]])

```

Nothing significantly changed in fiber group (n = `r as.character(n_fiber)`). 

## PCA

```{r}
fermented_df <- full_fermented_data %>% spread(key = Analyte, value = value) %>% dplyr::filter(Timepoint == 1 | Timepoint == 6)

fermented_pca <- fermented_df %>% select(-Participant, -Timepoint) %>% prcomp(.)

fermented_pca_df <- data.frame(Participant = fermented_df$Participant, 
                               Timepoint = as.factor(fermented_df$Timepoint), 
                               PC1 = fermented_pca$x[,"PC1"],
                               PC2 = fermented_pca$x[,"PC2"])

ggplot(fermented_pca_df, aes(x = PC1, y = PC2)) + geom_point(aes(colour = Timepoint))
```


