---
title: "16s_food_pilot"
author: "HCW"
date: "2/15/2019"
output: html_document
---

```{r include=FALSE}
library(ggplot2)
library(dplyr)
library(phyloseq)
library(dada2)
```

Import all the raw sequence tables (OTU tables)
```{r}
save_path <- "../../data/16S/"
save_path_figures <- "../../plots/"
data_path <- "../../data/16S/"

#food phyloseq object
food_psobj <- readRDS(file = paste(save_path, "fermented_food_ps_obj_fwdOnly.rds", sep = "")) 

food_seqtab <- food_psobj@otu_table %>% t() %>% as.data.frame() 
food_seqtab <- food_seqtab[,colSums(food_seqtab)>0]
food_samdf <- data.frame(food_psobj@sam_data) %>% 
  mutate(SampleID_merge=paste("food",SampleID,sep="_")) %>% 
  select(-c(BarcodeSequence:templatewell),-c(date:location))
rownames(food_seqtab) <- food_samdf$SampleID_merge

food_seq_df <- cbind(food_samdf,food_seqtab) 
food_seq_df_gather <- food_seq_df %>% 
  gather(key=ASV,value=ASV_count,-c(SampleID:SampleID_merge))

ASV_label_key <- food_seq_df_gather %>% 
  select(ASV) %>% 
  unique %>% 
  mutate(ASV_label=paste("food_ASV",1:dim(.)[1],sep="_"))

food_taxatab <- food_psobj@tax_table %>% 
  as.data.frame() %>% 
  mutate(ASV=rownames(.)) %>% 
  left_join(ASV_label_key,.) %>% 
  left_join(food_seq_df_gather,.)

food_taxatab_filt <- food_taxatab %>% 
  filter(ASV_count>250)

#collapse duplicates in taxa table (sum) 
food_taxatab_filt_taxa_collapse <- food_taxatab_filt %>% 
  group_by(type,Kingdom,Phylum,Class,Order,Family,Genus) %>% 
  dplyr::summarize(ASV_count_sum=sum(ASV_count))

food_taxatab_filt_taxa_collapse_spread <- food_taxatab_filt_taxa_collapse %>% 
  spread(key=type,value=ASV_count_sum) %>% 
  mutate(Genus_labeled=ifelse(Genus %in% c("g__",NA),paste(Order,Family,sep = "_"),as.character(Genus)))
food_taxatab_filt_taxa_collapse_spread[is.na(food_taxatab_filt_taxa_collapse_spread)] <- 0
#write.csv(food_taxatab_filt_taxa_collapse_spread,paste(save_path_figures,"fermented_food_taxa_table.csv"))

#stack barplot of taxa in food
food_taxatab_filt_taxa_collapse_gather <- food_taxatab_filt_taxa_collapse_spread %>% 
  select(-Genus) %>% 
  gather(key=fermented_food_type,value=ASV_count_sum,-Kingdom,-Phylum,-Class,-Order,-Family,-Genus_labeled)

nb.cols <- 16
mycolors <- colorRampPalette(brewer.pal(8, "Set3"))(nb.cols)
ggplot(food_taxatab_filt_taxa_collapse_gather, aes(fill=Genus_labeled, y=ASV_count_sum, x=fermented_food_type)) + 
    geom_bar(position="fill", stat="identity")+
  theme_bw()+
  scale_fill_manual(values = mycolors)+
  theme(text = element_text(size=12))
ggsave(paste(save_path_figures,"fermented_foods_stacked_barplot.pdf"),height=8,width=10,useDingbats=FALSE)

#participant phyloseq object
fermented_ppt_psobj <- readRDS( file = paste(save_path, "phyloseq_obj_PilotStudy_tree_redo_fwdONLY.rds", sep = "")) %>% 
  subset_samples(Group=="Fermented")

ppt_seqtab <- fermented_ppt_psobj@otu_table %>% as.data.frame() 
ppt_seqtab <- ppt_seqtab[,colSums(ppt_seqtab!=0)]

ppt_samdf <- data.frame(fermented_ppt_psobj@sam_data) %>%
  mutate(SampleID_merge=paste("ppt",SampleID,sep="_")) %>% 
  select(-c(BarcodeSequence:Stool_Kit_ID),-Description) 
rownames(ppt_seqtab) <- ppt_samdf$SampleID_merge
#None of the sig. ASVs changed in the fermented food cohort are ASVs found in the fermented food itself
# df_ranks_fermented_gather_filt_FOOD <- df_ranks_fermented_gather_filt %>% 
#   filter(ASV %in% c(food_asvs))
```


Merge the two sequence tables using the function mergeSequenceTables
- First want to add a prefix to the sequence table sample names
- For the mergeSequenceTable(table1,table2), rownames correspond to samples and column names correspond to sequences
```{r}
seqtab_all <- mergeSequenceTables(otu_table(food_seqtab,taxa_are_rows=F),otu_table(ppt_seqtab,taxa_are_rows=F)) %>% as.data.frame(.)
```


```{r}
food_asvs <- subset(seqtab_all, rownames(seqtab_all) %in% rownames(food_seqtab))
food_asv_list <- colnames(food_asvs[,colSums(food_asvs)>0])
ppt_asvs <- subset(seqtab_all, rownames(seqtab_all) %in% rownames(ppt_seqtab))
ppt_asvs_list <- colnames(ppt_asvs[,colSums(ppt_asvs)>0])

ppt_list <- unique(ppt_samdf$Participant) 

total_new=length(ppt_list)
new_in_food=length(ppt_list)
for(i in 1:length(ppt_list)){
  ppt_base_df = subset(ppt_asvs,rownames(ppt_asvs) %in% filter(ppt_samdf,Participant==ppt_list[i] & Timepoint %in% c(1,2))$SampleID_merge)
  ppt_base_zero = colnames(ppt_base_df[,colSums(ppt_base_df)==0])
  ppt_maint_df = subset(ppt_asvs,rownames(ppt_asvs) %in% filter(ppt_samdf,Participant==ppt_list[i] & Timepoint %in% c(3:9))$SampleID_merge)
  ppt_main_nonzero = colnames(ppt_maint_df[,colSums(ppt_maint_df)>0])
  ppt_gained = intersect(ppt_base_zero,ppt_main_nonzero)
  total_new[i]=length(ppt_gained)
  ppt_gained_food = intersect(ppt_gained,food_asv_list)
  new_in_food[i]=length(ppt_gained_food)
}

food_ppt_asvs_df <- data.frame(Participant=ppt_list,
                               total_new_asvs=total_new,
                               new_asvs_in_food=new_in_food,
                               percen_from_food=new_in_food/total_new)
```

Calculate total percentage of new ASVs from fermented food from the entire group
```{r}
total_new_fermented_group <- sum(food_ppt_asvs_df$total_new_asvs)
total_new_food_asvs_fermented_group <- sum(food_ppt_asvs_df$new_asvs_in_food)

total_new_food_asvs_fermented_group/total_new_fermented_group

pie_chart_df <- data.frame(Label=c("Found in fermented food",
                                   "Not found in fermented food"),
                           Percent=c(total_new_food_asvs_fermented_group/total_new_fermented_group,
                                     1-total_new_food_asvs_fermented_group/total_new_fermented_group))
ggplot(pie_chart_df, aes(x="", y=Percent, fill=Label)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void()+
  scale_fill_manual(breaks = c("Found in fermented food", "Not found in fermented food"), 
                       values=c("#bebada", "#d9d9d9"))+
  labs(fill = "Percent new ASVs")
#ggsave(paste(save_path_figures,"fermented_pie_chart_food_asvs.pdf"),width = 4.5,height=4,useDingbats=FALSE)

```

